Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-14 appendix.tex
more explicit (and hopefully correct) likelihood form

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2,w\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
%EM It seems to me that this only works if we know that \tilde{t} = t^*$. Otherwise the divergence won't have this form.
%EM Also, I tried to make the connection between the divergence and g a little more explicit.
The second derivative of the divergence term $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t}) = -g(x_1)\log g(x_1)$ is
\[
-g''(x_1)(1+\log g(x_1)) - \frac{g'(x_1)^2}{g(x_1)}.
\]
Using each functional form in Table~\ref{tab:gen-sitepatprob} we see in all cases $g''(x_1)=0$ and $g(x_1) > 0$ for $x_1\in(0,1).$
This implies the above second derivative is negative for each component of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$.

We can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ using Table~\ref{tab:farris_likelihoods}.
%EM This falls short of being "clearly"--- I can easily see that we get some log(1+x_1) terms and some log(1-x_1) terms, and that the below follows if the sum of the coefficients for the two terms is < 1. Perhaps add a sentence here?
Clearly this partial log-likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in(0,1)$ and thus has a unique maximum.
Since both summands of the likelihood are concave on $x_1\in(0,1)$, the likelihood is as well, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

>>>>

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
%EM It seems to me that this only works if we know that \tilde{t} = t^*$. Otherwise the divergence won't have this form.
%das: Heh, yeah oops, that was super sloppy.
%Should be fixed, and a lot more explicit now; thanks for keeping me honest!
%Also there's no restriction on w being not equal to zero or one since we're only interested in x and y, so that was removed, too.
%EM Also, I tried to make the connection between the divergence and g a little more explicit.
Define
\[
g_j(x_1) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}),
\]
i.e., the generating probability of the $j$th site split as a function of $x_1$.
Recall the divergence
\[
\shannonDivergence_{\tau^*,t^*}(\tau,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \tilde{t}),
\]
which, suppressing notation for clarity, we define as a function of $x_1$ as
\[
D(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \log g_j(x_1)
\]
where $c_j \ge 0$ for $j=1,\ldots,\nSiteSplits$.
It remains to show $D''(x_1) < 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
By straightforward calculus,
\[
D''(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j(x_1)g_j''(x_1) - g_j'(x_1)^2}{g_j(x_1)^2}\right).
\]
Using each functional form in Table~\ref{tab:gen-sitepatprob} we see $g_j''(x_1)=0$ and $g_j(x_1)>0$ for $x_1,y_1,x_2,y_2\in(0,1)$ and $j=1,\ldots,\nSiteSplits$ so that
\[
D''(x_1) = -\sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j'(x_1)}{g_j(x_1)}\right)^2.
\]
Finally, we require at least one $j$ where $c_j \neq 0$ and $g_j'(x_1) \neq 0$ so that $D''(x_1) \not\equiv 0$.
Consider the case $j=1$ where $\siteSplit_1=\emptyset$ so that, ignoring the $1/8$ constant in each term,
\[
g_1(x_1) \propto 1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2,
\]
\[
g_1'(x_1) \propto x_2 + y_1w + y_2w + y_1x_2y_2,
\]
and
\[
c_1 \propto 1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2.
\]
We have $c_1 > 0$ by $x^*,y^*\in[0,1]$ and, by assumption, $g_1'(x_1) > 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
This implies the above second derivative is negative.

%EM This falls short of being "clearly"--- I can easily see that we get some log(1+x_1) terms and some log(1-x_1) terms, and that the below follows if the sum of the coefficients for the two terms is < 1. Perhaps add a sentence here?
%das: you're right, I missed a factor of two; hope it's more clear now.
We now focus on the partial log-likelihood term.
Recall the partial log-likelihood takes the form
\[
\tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, \tilde{t}).
\]
As above, define the partial log-likelihood as a function of $x_1$ as
\[
\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_j(x_1)
\]
where
\[
h_j(x_1) = \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}).
\]
Using the procedure in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
\[
h_j(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
\end{align}
This has a second derivative
\[
\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that, by the second derivative test, the sum of $D(x_1)$ and $\tilde{\ell}(x_1)$ has a unique maximum.
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$, thus, $D(x_1) + \tilde{\ell}(x_1)$ is concave as a function of $x_1\in(0,1)$, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
